---
title: "Gender Differences in Behavioral Responses to Social Status"
author: "Senne Hollard"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    df_print: paged
    code_folding: show
subtitle: Empirical analysis — Study I (loss aversion in social image) and Study II
  (social status and prosocial behavior)
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 7,
  fig.height = 4.6,
  fig.align  = "center",
  dpi = 150
)

## Packages (base-R + a minimal, widely available set).
library(haven)      # read Stata .dta
library(ggplot2)    # figures
library(survival)   # survreg() -> two-limit Tobit
library(MASS)       # polr() ordered probit for Study II
library(knitr)      # kable() tables

set.seed(5315751)   # reproducibility (student number); used by the cluster bootstrap

## Number of session-cluster bootstrap replications used for Study I SEs.
B_BOOT <- 499
## Number of session-cluster bootstrap replications used for Study II ordered probit SEs.
B_OP <- 199
```

This document reproduces exactly the empirical analysis reported in the thesis
*Analysis of Gender Differences in Behavioral Responses to Experimentally
Manipulated Social Status*. It reanalyses two publicly available replication
packages through a gender-heterogeneity lens.

* **Study I** — Petrishcheva, Riener & Schildberg-Hörisch (2023), *Loss aversion
  in social image concerns*. A regression-discontinuity (RD) design in dishonest
  reporting around a rank-based reference point, estimated **separately for men,
  women, and the full sample** (thesis Eq. 3), plus a **pooled gender-interaction
  Tobit** (thesis Eq. 5).
* **Study II** — Zheng, Schram & Song (2023b), *Social status and prosocial
  behavior*. The original status-treatment model **extended with gender
  interactions** (thesis Eq. 7), with a **Wald test** of whether the high-status
  gender interaction differs between randomly assigned and earned status.

Deliberate, transparent modelling choices (kept minimal and reproducible):

1. **Clustered standard errors.** Study I uses a *session-level cluster
   bootstrap* (`r B_BOOT` replications) for the Tobit RD estimates and the pooled
   gender-difference Tobit. Study II uses a separate session-level cluster
   bootstrap (`r B_OP` replications) for the ordered probit coefficients.

---

# Setup: helper functions

All estimation relies on a few small, self-contained helpers.

```{r helpers}
## --- (a) Stata-labelled .dta columns -> plain numeric --------------------
to_numeric <- function(df) {
  as.data.frame(lapply(df, function(x)
    if (inherits(x, "haven_labelled")) as.numeric(x) else x))
}

## --- (b) Two-limit Tobit via survreg -------------------------------------
## Latent DiceDif* = X'b + e is observed censored to [lo, hi] = [-5, 5]
## (thesis Eq. 4). Coded as interval-censored data: 1 = exact, 2 = left-censored
## at lo, 0 = right-censored at hi.
fit_tobit <- function(formula, data, lo = -5, hi = 5) {
  mf <- model.frame(formula, data, na.action = na.omit)
  y  <- model.response(mf)
  event <- ifelse(y <= lo, 2L, ifelse(y >= hi, 0L, 1L))
  mf$.time  <- pmin(pmax(y, lo), hi)
  mf$.event <- event
  rhs <- paste(deparse(formula[[3]]), collapse = " ")
  f   <- as.formula(paste("Surv(.time, .time, .event, type = 'interval') ~", rhs))
  suppressWarnings(
    survreg(f, data = mf, dist = "gaussian",
            control = survreg.control(maxiter = 200))
  )
}

## --- (c) Session-level cluster bootstrap ----------------------------------
## Resamples whole clusters (sessions) with replacement and refits.
## `refit` must return a named coefficient vector (or NULL on failure).
cluster_boot_se <- function(data, cluster, refit, B = B_BOOT) {
  cl     <- data[[cluster]]
  groups <- unique(cl)
  est0   <- refit(data)
  reps   <- matrix(NA_real_, B, length(est0), dimnames = list(NULL, names(est0)))
  ok <- 0L
  for (b in seq_len(B)) {
    gb <- sample(groups, length(groups), replace = TRUE)
    db <- do.call(rbind, lapply(gb, function(g) data[cl == g, , drop = FALSE]))
    e  <- tryCatch(refit(db), error = function(err) NULL)
    if (!is.null(e) && length(e) == length(est0)) { ok <- ok + 1L; reps[ok, ] <- e }
  }
  reps <- reps[seq_len(ok), , drop = FALSE]
  list(se = apply(reps, 2, sd, na.rm = TRUE), nrep = ok)
}

## --- (d) Ordered-outcome and ordered-probit bootstrap helpers -------------
## Keeps the observed numeric category order but passes an ordered factor to polr().
ordered_numeric_outcome <- function(x, levels = NULL) {
  if (is.null(levels)) levels <- sort(unique(x[!is.na(x)]))
  factor(x, levels = levels, ordered = TRUE)
}

## Resamples whole sessions, refits the ordered probit, and stores only
## regression coefficients; polr threshold/cutpoint parameters are excluded.
cluster_boot_op <- function(data, cluster, refit, coef_names, B = B_OP) {
  cl     <- data[[cluster]]
  groups <- unique(cl)
  reps   <- matrix(NA_real_, B, length(coef_names),
                   dimnames = list(NULL, coef_names))
  ok <- 0L
  for (b in seq_len(B)) {
    gb <- sample(groups, length(groups), replace = TRUE)
    db <- do.call(rbind, lapply(gb, function(g) data[cl == g, , drop = FALSE]))
    e  <- tryCatch(refit(db), error = function(err) NULL)
    if (!is.null(e) && all(coef_names %in% names(e))) {
      e <- e[coef_names]
      if (all(is.finite(e))) {
        ok <- ok + 1L
        reps[ok, ] <- e
      }
    }
  }
  reps <- reps[seq_len(ok), , drop = FALSE]
  V <- if (ok > 1L) stats::cov(reps) else
    matrix(NA_real_, length(coef_names), length(coef_names),
           dimnames = list(coef_names, coef_names))
  list(se = sqrt(diag(V)), vcov = V, reps = reps, nrep = ok)
}

## --- (e) Wald test of a single linear restriction L'b = 0 -----------------
## Given coefficients `b` and a covariance matrix `V`, tests a
## restriction supplied as a named vector, e.g. c("HRxF" = 1, "HExF" = -1) for
## H0: theta2 = theta4. Returns the contrast, its SE, the chi-square(1)
## statistic and the p-value (a chi^2(1) Wald test, as stated in the thesis).
wald_test <- function(b, V, restriction) {
  L <- setNames(numeric(length(b)), names(b))
  L[names(restriction)] <- restriction
  diff <- sum(L * b)
  vard <- as.numeric(t(L) %*% V %*% L)
  W    <- diff^2 / vard
  list(diff = diff, se = sqrt(vard), z = diff / sqrt(vard),
       chisq = W, df = 1L, p = pchisq(W, 1, lower.tail = FALSE))
}

## --- (f) Significance stars + cell formatting -----------------------------
stars  <- function(p) ifelse(is.na(p), "",
            ifelse(p < .01, "***", ifelse(p < .05, "**", ifelse(p < .1, "*", ""))))
fmt_b  <- function(b, p, d = 3) paste0(formatC(b,  format = "f", digits = d), stars(p))
fmt_se <- function(s, d = 3)   paste0("(", formatC(s, format = "f", digits = d), ")")

## --- (g) Approximate post-estimation power (two-sided, 5%) -----------------
Z95 <- qnorm(0.975); Z80 <- qnorm(0.80)
power_now <- function(b, s) pnorm(abs(b)/s - Z95) + pnorm(-abs(b)/s - Z95)
mde_80    <- function(s)    (Z95 + Z80) * s                 # min. detectable effect
nreq_80   <- function(b, s, n) n * ((Z95 + Z80) * s / abs(b))^2

power_row <- function(label, b, se, n) data.frame(
  Coefficient        = label,
  Estimate           = formatC(b,  format = "f", digits = 3),
  `Clustered SE`     = formatC(se, format = "f", digits = 3),
  `Current power`    = formatC(power_now(b, se), format = "f", digits = 2),
  `MDE (80% power)`  = formatC(mde_80(se),        format = "f", digits = 3),
  `Required N (80%)` = formatC(nreq_80(b, se, n), format = "f", digits = 0),
  check.names = FALSE, stringsAsFactors = FALSE)
```

---

# Data loading

```{r load-data}
loss_raw <- to_numeric(read_dta(
  "0Loss_aversion_data/Loss Aversion in Social Image Concerns No Raven- Replication package/data-analysis/ExperimentMain.dta"))
prosocial_raw <- to_numeric(read_dta("0Prosocial_behavior_data/Pooled_data_April27.dta"))
```

---

# Data cleaning

## Study I — RD analysis variables

`DiceDif` is the die-roll difference (`DiceSubject - DiceSample`, bounded in
`[-5, 5]`, thesis Eq. 2); `RankDif` (`= Rank1 - P2Rank`, thesis Eq. 1) is the
running variable; `Loss = 1` on the loss side of the cutoff `RankDif = 0`. The
gender split adds `Male`.

```{r clean-study1}
s1 <- loss_raw
s1$Male <- 1 - s1$Female

## Main local sample around the loss-gain threshold.
s1$in_window <- abs(s1$RankDif) <= 10

## Control vector X (field-of-study base = "other"); gender added only in the
## full-sample column.
S1_CTRL <- c("Age", "Age_sq", "econ", "psych",
             "Grade_Abi", "LossAversion", "SocialImage", "Risk", "P2Rank")

## Subsample helpers: full RankDif range (for the RD plots, as in the thesis
## figure) and the local estimation window |RankDif| <= 10 (for the models).
s1_by_gender <- function(which = c("full", "male", "female")) {
  which <- match.arg(which)
  d <- s1
  if (which == "male")   d <- d[d$Male == 1, ]
  if (which == "female") d <- d[d$Male == 0, ]
  d
}
s1_sample <- function(which = c("full", "male", "female")) {
  d <- s1_by_gender(match.arg(which))
  d[abs(d$RankDif) <= 10, ]
}

cat("Local window |RankDif| <= 10:",
    sum(s1$in_window), "subjects",
    "(male", sum(s1$in_window & s1$Male == 1),
    "/ female", sum(s1$in_window & s1$Male == 0), ")\n")
```

## Study II — interaction-model variables

Player-2 observations only. Status dummies `RSLow / RSHigh / ESLow / ESHigh`
(low/high status under random/earned assignment) with the no-status baseline as
the omitted category; `Female`, `NL` (Netherlands), `age`, `econ` as controls;
sessions identified by `uniquesession`.

```{r clean-study2}
s2 <- prosocial_raw[prosocial_raw$player == 2 & !is.na(prosocial_raw$player), ]

s2$Female  <- s2$gender          # 1 = female, 0 = male
s2$NL      <- s2$country         # 1 = Netherlands, 0 = China
s2$cluster <- s2$uniquesession

## Condition label for figures / grouping.
s2$condition <- factor(s2$type, levels = 1:5,
  labels = c("Baseline", "Random low", "Random high", "Earned low", "Earned high"))
s2$Gender <- factor(s2$Female, levels = c(0, 1), labels = c("Male", "Female"))

cat("Player-2 observations:", nrow(s2), " | sessions:", length(unique(s2$cluster)), "\n")
print(table(Condition = s2$condition))
```

---

# Study I: Loss Aversion in Social Image Concerns

We keep the original RD design and split the sample by gender. The local Tobit
RD specification (thesis Eq. 3) is

$$DieDiff^{*} = \alpha + \tau\,Loss + \beta_1 RankDif + \beta_2 (Loss\times RankDif) + X'\gamma + u,$$

with `DiceDif` censored to `[-5, 5]` (thesis Eq. 4) and estimated on the local
sample `|RankDif| <= 10`. The coefficient of interest is $\tau$ (the coefficient
on `Loss`): the jump in over-reporting when crossing from small gains to small
losses. We report it **without** and **with** individual controls. All standard
errors are session-level cluster bootstrap (`r B_BOOT` replications).

## RD estimation

```{r study1-estimate}
## RHS strings for the Tobit RD (Eq. 3), with/without controls.
rhs_base <- "Loss + RankDif + Loss:RankDif"
rhs_ctrl <- function(add_female)
  paste(rhs_base,
        if (add_female) "+ Female" else "",
        "+", paste(S1_CTRL, collapse = " + "))

## Estimate tau + bootstrap SE for one Tobit specification on one sample.
est_tau <- function(sample, type = c("tobit_nc", "tobit_c")) {
  type <- match.arg(type)
  d    <- s1_sample(sample)
  add_female <- sample == "full"
  rhs  <- if (type == "tobit_nc") rhs_base else rhs_ctrl(add_female)
  fm   <- as.formula(paste("DiceDif ~", rhs))

  m  <- fit_tobit(fm, d)
  b  <- coef(m)[["Loss"]]
  N  <- length(m$linear.predictors)
  coefer <- function(dd) tryCatch(coef(fit_tobit(fm, dd)), error = function(e) NULL)
  bs <- cluster_boot_se(d, "Session", coefer, B = B_BOOT)
  se <- bs$se[["Loss"]]
  z  <- b / se
  list(b = b, se = se, p = 2 * pnorm(-abs(z)), N = N, nrep = bs$nrep)
}

samples    <- c("male", "female", "full")
estimators <- c("tobit_nc", "tobit_c")
set.seed(5315751)                          # reproducible RD-table bootstrap
s1_res <- setNames(lapply(estimators, function(ty)
            setNames(lapply(samples, est_tau, type = ty), samples)), estimators)

## Pooled gender-interaction Tobit (thesis Eq. 5): the formal test of the gender
## difference in the discontinuity is the coefficient on Loss:Male.
fm_pool   <- DiceDif ~ Loss * Male + RankDif + Loss:RankDif
pool_fit  <- fit_tobit(fm_pool, s1_sample("full"))
pool_coef <- function(dd) tryCatch(coef(fit_tobit(fm_pool, dd)), error = function(e) NULL)
set.seed(5315751)                          # reproducible pooled bootstrap
pool_bs   <- cluster_boot_se(s1_sample("full"), "Session", pool_coef, B = B_BOOT)
s1_gender_diff <- list(
  b  = coef(pool_fit)[["Loss:Male"]],
  se = pool_bs$se[["Loss:Male"]],
  N  = length(pool_fit$linear.predictors)
)
s1_gender_diff$p <- 2 * pnorm(-abs(s1_gender_diff$b / s1_gender_diff$se))
```

## RD figures (thesis Figure 1)

Each plot shows `DiceDif` (y) against `RankDif` (x), with binned means ($\approx$ 20
equal-frequency bins, as in the thesis figure), separate linear fits on the
loss and gain sides, a solid line at the cutoff `RankDif = 0`, and dashed lines
marking the `|RankDif| <= 10` local window.

```{r study1-plots, fig.width=7, fig.height=4.6}
bin_means <- function(df, nbin = 10) {
  do.call(rbind, lapply(split(df, df$Loss), function(g) {
    br <- unique(quantile(g$RankDif, seq(0, 1, length.out = nbin + 1), na.rm = TRUE))
    if (length(br) < 3)
      return(data.frame(RankDif = mean(g$RankDif), DiceDif = mean(g$DiceDif),
                        Loss = g$Loss[1]))
    g$.bin <- cut(g$RankDif, br, include.lowest = TRUE)
    a <- aggregate(cbind(RankDif, DiceDif) ~ .bin, g, mean)
    data.frame(RankDif = a$RankDif, DiceDif = a$DiceDif, Loss = g$Loss[1])
  }))
}

rd_plot <- function(which, title) {
  d <- s1_by_gender(which)          # full range, as in the thesis figure
  d$grp <- factor(d$Loss, c(0, 1), c("Gain", "Loss"))
  pts <- bin_means(d); pts$grp <- factor(pts$Loss, c(0, 1), c("Gain", "Loss"))
  ggplot(d, aes(RankDif, DiceDif, color = grp)) +
    geom_vline(xintercept = c(-10, 10), linetype = "dashed", color = "grey75") +
    geom_vline(xintercept = 0, color = "grey30") +
    geom_smooth(method = "lm", formula = y ~ x, se = FALSE, linewidth = 0.8) +
    geom_point(data = pts, aes(shape = grp), size = 2.3) +
    scale_color_manual(values = c(Gain = "#2c6fbb", Loss = "#b5402f")) +
    scale_shape_manual(values = c(Gain = 16, Loss = 18)) +
    coord_cartesian(xlim = c(-60, 60)) +
    labs(title = title, x = "Rank difference (RankDif)",
         y = "Die roll difference (DiceDif)", color = NULL, shape = NULL) +
    theme_minimal(base_size = 12) +
    theme(legend.position = "bottom",
          plot.title = element_text(face = "bold", size = 12))
}

p_rd_male   <- rd_plot("male",   "(a) Male sample")
p_rd_female <- rd_plot("female", "(b) Female sample")
p_rd_full   <- rd_plot("full",   "(c) Full sample")

p_rd_male
p_rd_female
p_rd_full
```

## RD table (thesis Table 5)

Columns are **Male / Female / Full**; Panel A is without controls, Panel B adds
the full control vector $X$. Both panels are two-limit Tobit models (Eq. 3).

```{r study1-table}
panel_rows <- function(res, label) {
  cell  <- function(s) c(fmt_b(s$b, s$p), fmt_se(s$se),
                         formatC(s$p, format = "f", digits = 3),
                         formatC(s$N, format = "d"))
  M <- cbind(cell(res$male), cell(res$female), cell(res$full))
  data.frame(
    Statistic = c(paste0(label, ": RD estimate (tau)"),
                  "  Cluster-bootstrap SE", "  p-value", "  N"),
    Male = M[, 1], Female = M[, 2], Full = M[, 3],
    check.names = FALSE, stringsAsFactors = FALSE)
}

s1_table <- rbind(
  panel_rows(s1_res$tobit_nc, "Panel A: Tobit RD, no controls"),
  panel_rows(s1_res$tobit_c,  "Panel B: Tobit RD, with controls")
)
rownames(s1_table) <- NULL

kable(s1_table, align = c("l", "r", "r", "r"),
      caption = paste0("Study I - gender-split two-limit Tobit RD estimates of ",
        "the loss/gain discontinuity in over-reporting (tau = coefficient on ",
        "Loss, Eq. 3). Local sample |RankDif| <= 10. Session cluster-bootstrap ",
        "SEs (", B_BOOT, " reps). *** p<.01, ** p<.05, * p<.1."))
```

## Pooled gender-difference test (thesis Eq. 5)

```{r study1-genderdiff, results='asis'}
cat(sprintf(paste0(
  "\n**Pooled gender-difference test (Eq. 5).** Coefficient on `Loss x Male` ",
  "in a full-sample Tobit = **%.3f** (cluster-bootstrap SE %.3f, p = %.3f, ",
  "N = %d). A negative value means the loss/gain discontinuity is *smaller* for ",
  "men than for women.\n"),
  s1_gender_diff$b, s1_gender_diff$se, s1_gender_diff$p, s1_gender_diff$N))
```

## Power diagnostics (thesis Table 6)

Approximate post-estimation power for the pooled gender-difference coefficient
`Loss x Male`, which directly captures the male–female difference in the
discontinuity.

```{r study1-power}
s1_power <- power_row("Loss x Male", s1_gender_diff$b, s1_gender_diff$se,
                      s1_gender_diff$N)
kable(s1_power, align = c("l", rep("r", 5)),
      caption = paste0("Study I - approximate post-estimation power diagnostics ",
        "for the pooled gender difference, full-sample Tobit on the local window ",
        "(N = ", s1_gender_diff$N, "). Approximate diagnostics, not exact ",
        "ex-ante calculations."))
```

---

# Study II: Social Status and Prosocial Behavior

The original status-treatment model is extended with a `Female` main effect and
four status × `Female` interactions (thesis Eq. 7). The ordered-probit latent
index is

$$Y^* = \alpha + \beta_1 LR + \beta_2 HR + \beta_3 LE + \beta_4 HE + \delta\,Female
      + \textstyle\sum_k \theta_k (\text{status}_k \times Female) + Z'\gamma + \varepsilon,$$

with controls $Z$ = Netherlands, age, econ major. Standard errors come from a
session-level cluster bootstrap. **Chosen effort** is the primary outcome;
**mean response** and **proposal received** are reported as secondary ordered
outcomes. Here $\theta_2$ is the coefficient on $HR\times Female$ and
$\theta_4$ that on $HE\times Female$. A positive coefficient indicates a higher
latent propensity to choose a higher outcome category; a negative interaction
indicates that the corresponding status effect is smaller for women than for men
on the ordered-probit latent index.

## Interaction-model estimation

```{r study2-estimate}
rhs2 <- paste("RSLow + RSHigh + ESLow + ESHigh + Female",
              "+ RSLow:Female + RSHigh:Female + ESLow:Female + ESHigh:Female",
              "+ NL + age + econ")

S2_RHS_VARS <- c("RSLow", "RSHigh", "ESLow", "ESHigh", "Female",
                 "NL", "age", "econ")

fit_s2 <- function(y, label) {
  vars <- c(y, S2_RHS_VARS, "cluster")
  d <- s2[complete.cases(s2[, vars]), ]
  y_levels <- sort(unique(d[[y]][!is.na(d[[y]])]))

  if (length(y_levels) < 3L) {
    return(list(ok = FALSE, outcome = label,
                reason = "Fewer than three ordered outcome categories."))
  }

  d$.y_ord <- ordered_numeric_outcome(d[[y]], y_levels)
  fm <- as.formula(paste(".y_ord ~", rhs2))
  m <- tryCatch(
    MASS::polr(fm, data = d, method = "probit", Hess = TRUE, model = TRUE),
    error = function(e) e
  )

  if (inherits(m, "error")) {
    return(list(ok = FALSE, outcome = label, reason = conditionMessage(m)))
  }

  coef_names <- names(coef(m))
  refit <- function(dd) {
    dd$.y_ord <- ordered_numeric_outcome(dd[[y]], y_levels)
    mm <- MASS::polr(fm, data = dd, method = "probit", Hess = TRUE, model = TRUE)
    coef(mm)[coef_names]
  }

  bs <- cluster_boot_op(d, "cluster", refit, coef_names, B = B_OP)
  list(ok = TRUE, outcome = label, model = m, data = d, y = y,
       y_levels = y_levels, vcov = bs$vcov, boot = bs,
       n = nobs(m), n_clusters = length(unique(d$cluster)))
}

set.seed(5315751)                          # reproducible Study II bootstrap
s2_cho  <- fit_s2("choseneffort2", "Chosen effort")      # primary
s2_mean <- fit_s2("meanresponse",  "Mean response")      # secondary
s2_prop <- fit_s2("proposal",      "Proposal received")  # secondary

s2_models <- list(`Chosen effort` = s2_cho,
                  `Mean response` = s2_mean,
                  `Proposal received` = s2_prop)
s2_ok <- vapply(s2_models, function(x) isTRUE(x$ok), logical(1))
if (any(!s2_ok)) {
  cat("Ordered probit not estimated for:",
      paste(names(s2_models)[!s2_ok], collapse = ", "), "\n")
  cat("Reason(s):",
      paste(vapply(s2_models[!s2_ok], `[[`, character(1), "reason"),
            collapse = " | "), "\n")
}
if (!isTRUE(s2_cho$ok))
  stop("Chosen-effort ordered probit failed; no OLS fallback is used.")
s2_models_ok <- s2_models[s2_ok]

## Tidy coefficient + bootstrap-SE extractor for ordered probit.
tidy_op <- function(f) {
  b  <- coef(f$model)
  se <- f$boot$se[names(b)]
  z  <- b / se
  data.frame(term = names(b), b = b, se = se, z = z,
             p = 2 * pnorm(-abs(z)),
             row.names = NULL, stringsAsFactors = FALSE)
}
```

## Chosen-effort figure (thesis Figure 2)

Chosen effort (y) against proposed effort (x), with **separate male and female
lines** in **five panels**, one per status condition. Values are the
strategy-method responses `effort{1,3,5,7,9}AS` averaged within
gender × condition × proposal.

```{r study2-plot, fig.width=8.5, fig.height=5.4}
prop_levels <- c(1, 3, 5, 7, 9)
s2_long <- do.call(rbind, lapply(prop_levels, function(j)
  data.frame(condition = s2$condition, Gender = s2$Gender,
             proposed = j, response = s2[[paste0("effort", j, "AS")]])))

s2_curve <- aggregate(response ~ condition + Gender + proposed, s2_long, mean)

p_s2 <- ggplot(s2_curve, aes(proposed, response, color = Gender, group = Gender)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  facet_wrap(~ condition, nrow = 2) +
  scale_x_continuous(breaks = prop_levels) +
  scale_color_manual(values = c(Male = "#2c6fbb", Female = "#b5402f")) +
  labs(x = "Proposed effort", y = "Chosen effort (mean response)", color = NULL,
       title = "Conditional chosen effort by gender and status condition") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom",
        plot.title = element_text(face = "bold", size = 12))
p_s2
```

## Regression table (thesis Table 7)

```{r study2-table}
S2_TERMS <- c("RSLow", "RSHigh", "ESLow", "ESHigh", "Female",
              "RSLow:Female", "RSHigh:Female", "ESLow:Female", "ESHigh:Female",
              "NL", "age", "econ")
S2_LABELS <- c("Low random status (LR)", "High random status (HR)",
               "Low earned status (LE)", "High earned status (HE)", "Female",
               "LR x Female", "HR x Female", "LE x Female", "HE x Female",
               "Netherlands", "Age", "Econ major")

reg_column <- function(f) {
  td <- tidy_op(f); rownames(td) <- td$term
  out <- character(2 * length(S2_TERMS))
  for (i in seq_along(S2_TERMS)) {
    t <- S2_TERMS[i]
    out[2 * i - 1] <- if (t %in% td$term) fmt_b(td[t, "b"], td[t, "p"]) else ""
    out[2 * i]     <- if (t %in% td$term) fmt_se(td[t, "se"]) else ""
  }
  out
}

s2_table <- data.frame(Term = as.vector(rbind(S2_LABELS, "")),
                       check.names = FALSE, stringsAsFactors = FALSE)
for (nm in names(s2_models_ok)) s2_table[[nm]] <- reg_column(s2_models_ok[[nm]])

## Footer: N, clusters, and successful ordered-probit bootstrap replications.
foot <- data.frame(Term = c("N", "Sessions (clusters)",
                            "Successful bootstrap replications"),
                   check.names = FALSE, stringsAsFactors = FALSE)
for (nm in names(s2_models_ok)) {
  foot[[nm]] <- c(s2_models_ok[[nm]]$n,
                  s2_models_ok[[nm]]$n_clusters,
                  s2_models_ok[[nm]]$boot$nrep)
}
s2_table_full <- rbind(s2_table, foot)

kable(s2_table_full, align = c("l", rep("r", length(s2_models_ok))),
      caption = paste0("Study II - status treatment effects with gender ",
        "interactions (ordered probit, Eq. 7; player-2 sample). Coefficients ",
        "are latent-index ordered-probit coefficients, not effort-unit effects. ",
        "Session-cluster bootstrap SEs in parentheses. Baseline / no-status is ",
        "the omitted category. Status-type coefficients give the effect for men; ",
        "interaction terms the additional effect for women. *** p<.01, ** p<.05, ",
        "* p<.1."))
```

## Wald test: does the high-status gender interaction differ by source? (H2b)

H2b asks whether the gender difference in the high-status effect differs between
randomly assigned and earned status — i.e. whether $\theta_2 = \theta_4$, where
$\theta_2$ is the coefficient on $HR\times Female$ and $\theta_4$ that on
$HE\times Female$. The thesis tests this restriction with a **Wald test** in the
chosen-effort ordered-probit model, using the bootstrap covariance matrix so
that the test accounts for the covariance between the two interaction
coefficients on the latent index.

```{r study2-wald}
## H0: theta2 = theta4  <=>  (RSHigh:Female) - (ESHigh:Female) = 0.
w_cho <- wald_test(coef(s2_cho$model), s2_cho$vcov,
                   c("RSHigh:Female" = 1, "ESHigh:Female" = -1))

cho_td <- tidy_op(s2_cho); rownames(cho_td) <- cho_td$term
th2 <- cho_td["RSHigh:Female", "b"]   # theta2 (HR x Female)
th4 <- cho_td["ESHigh:Female", "b"]   # theta4 (HE x Female)

wald_tbl <- data.frame(
  Quantity = c("theta2  (HR x Female)", "theta4  (HE x Female)",
               "Difference  theta2 - theta4", "SE of difference",
               "Chi-square(1) statistic", "p-value"),
  Value = c(formatC(th2,         format = "f", digits = 3),
            formatC(th4,         format = "f", digits = 3),
            formatC(w_cho$diff,  format = "f", digits = 3),
            formatC(w_cho$se,    format = "f", digits = 3),
            formatC(w_cho$chisq, format = "f", digits = 3),
            formatC(w_cho$p,     format = "f", digits = 3)),
  check.names = FALSE, stringsAsFactors = FALSE)

kable(wald_tbl, align = c("l", "r"),
      caption = paste0("Study II - Wald test of H0: theta2 = theta4 ",
        "(equality of the high-random-status and high-earned-status gender ",
        "interactions) in the chosen-effort ordered-probit model. The test uses ",
        "the bootstrap covariance matrix for latent-index coefficients."))
```

### Copy-paste LaTeX (Wald-test paragraphs)

The two paragraphs below reproduce the thesis text with the `XX` placeholders
filled in from the numbers just computed. They are written verbatim and can be
pasted straight into the LaTeX source.

```{r study2-wald-latex, comment=""}
mean_td <- if (isTRUE(s2_mean$ok)) {
  x <- tidy_op(s2_mean); rownames(x) <- x$term; x
} else NULL

para1 <- sprintf(paste0(
"The estimates provide weaker evidence for \\textbf{H2b}. The gender difference in the high-status effect appears larger under random status than under earned status on the ordered-probit latent index: for chosen effort, the interaction is $%.3f$ for high random status but only $%.3f$ for high earned status; for mean response, the corresponding estimates are $%.3f$ and $%.3f$. To test this comparison directly rather than by inspection, I conduct a Wald test of the coefficient restriction $\\theta_2 = \\theta_4$ in the chosen-effort ordered-probit model, using the session-cluster bootstrap covariance matrix so that the test accounts for the covariance between the two interaction coefficients. The restriction cannot be rejected ($\\chi^2(1) = %.2f$, $p = %.2f$): although the high-random-status interaction is larger in magnitude than the high-earned-status interaction, the two latent-index coefficients are not statistically distinguishable from each other. The directional pattern is therefore consistent with the idea that gender differences are stronger when status is randomly assigned rather than earned, but the data cannot establish that the source of status significantly changes the gender difference. Overall, the results suggest that gender heterogeneity is concentrated in the random-status condition, while remaining too imprecise to confirm that random and earned status differ in this respect."),
  cho_td["RSHigh:Female", "b"], cho_td["ESHigh:Female", "b"],
  if (!is.null(mean_td)) mean_td["RSHigh:Female", "b"] else NA_real_,
  if (!is.null(mean_td)) mean_td["ESHigh:Female", "b"] else NA_real_,
  w_cho$chisq, w_cho$p)

para2 <- sprintf(paste0(
"The results provide suggestive but underpowered support for H2b: the gender difference is larger under randomly assigned status and smaller under earned status on the ordered-probit latent index, consistent with the prediction that random status makes the difference more pronounced, but a Wald test cannot reject equality of the two interactions ($p = %.2f$) and most interaction terms are imprecisely estimated."),
  w_cho$p)

cat("===== Paragraph 1 (Section 5.2.2, ordered-probit latent-index wording) =====\n\n")
cat(para1, "\n\n")
cat("===== Paragraph 2 (Section 5.2.3, ordered-probit latent-index wording) =====\n\n")
cat(para2, "\n")
```

## Latent-index implied status effects

The implied status effects below keep the original addition logic for main
effects and gender interactions, but they are interpreted only on the
ordered-probit latent index. They are not effort-point changes.

```{r study2-implied-effects}
S2_STATUS_TERMS <- data.frame(
  Status = c("Low random status", "High random status",
             "Low earned status", "High earned status"),
  main = c("RSLow", "RSHigh", "ESLow", "ESHigh"),
  interaction = c("RSLow:Female", "RSHigh:Female",
                  "ESLow:Female", "ESHigh:Female"),
  stringsAsFactors = FALSE
)

latent_effect_row <- function(f, status_label, main_term, int_term, gender) {
  b <- coef(f$model)
  L <- setNames(numeric(length(b)), names(b))
  L[main_term] <- 1
  if (gender == "Female") L[int_term] <- 1
  est <- sum(L * b)
  se <- sqrt(as.numeric(t(L) %*% f$vcov %*% L))
  z <- est / se
  data.frame(
    Gender = gender,
    Status = paste0(status_label, " - baseline"),
    Estimate = formatC(est, format = "f", digits = 3),
    `Bootstrap SE` = formatC(se, format = "f", digits = 3),
    `p-value` = formatC(2 * pnorm(-abs(z)), format = "f", digits = 3),
    check.names = FALSE, stringsAsFactors = FALSE)
}

s2_implied_effects <- do.call(rbind, lapply(seq_len(nrow(S2_STATUS_TERMS)), function(i) {
  r <- S2_STATUS_TERMS[i, ]
  rbind(latent_effect_row(s2_cho, r$Status, r$main, r$interaction, "Male"),
        latent_effect_row(s2_cho, r$Status, r$main, r$interaction, "Female"))
}))
rownames(s2_implied_effects) <- NULL

kable(s2_implied_effects, align = c("l", "l", "r", "r", "r"),
      caption = paste0("Study II - implied status effects in the chosen-effort ",
        "ordered-probit model. Estimates combine status main effects and, for ",
        "women, the corresponding status-by-Female interaction. These are ",
        "latent-index effects, not effort-point changes."))
```

## Predicted chosen effort

For interpretability, the table below reports model-based predicted expected
chosen effort under selected counterfactual status conditions, averaging over
the observed distribution of controls. These are predictions from the
chosen-effort ordered-probit model.

```{r study2-predicted-effort}
expected_ordered_outcome <- function(f, newdata) {
  probs <- predict(f$model, newdata = newdata, type = "probs")
  if (is.vector(probs)) probs <- matrix(probs, nrow = 1)
  as.numeric(probs %*% f$y_levels)
}

predict_expected_effort <- function(female, status_var) {
  nd <- s2_cho$data
  nd[, c("RSLow", "RSHigh", "ESLow", "ESHigh")] <- 0
  if (!is.na(status_var)) nd[[status_var]] <- 1
  nd$Female <- female
  mean(expected_ordered_outcome(s2_cho, nd))
}

pred_row <- function(gender, female) {
  baseline <- predict_expected_effort(female, NA_character_)
  hr <- predict_expected_effort(female, "RSHigh")
  he <- predict_expected_effort(female, "ESHigh")
  data.frame(
    Gender = gender,
    Baseline = formatC(baseline, format = "f", digits = 3),
    `High random status` = formatC(hr, format = "f", digits = 3),
    `High earned status` = formatC(he, format = "f", digits = 3),
    `High random - baseline` = formatC(hr - baseline, format = "f", digits = 3),
    `High earned - baseline` = formatC(he - baseline, format = "f", digits = 3),
    check.names = FALSE, stringsAsFactors = FALSE)
}

s2_predicted_effort <- rbind(pred_row("Male", 0), pred_row("Female", 1))

kable(s2_predicted_effort, align = c("l", rep("r", 5)),
      caption = paste0("Study II - model-based predicted expected chosen effort ",
        "from the ordered-probit model, averaged over observed controls. ",
        "Differences are on the expected chosen-effort scale."))
```

## Power diagnostics (thesis Table 8)

Approximate post-estimation power for the gender coefficients in the
chosen-effort ordered-probit model. These are latent-index diagnostics based on
ordered-probit coefficients and session-cluster bootstrap SEs; they are not
directly comparable to the previous OLS effort-point diagnostics.

```{r study2-power}
N_s2 <- s2_cho$n
power_terms <- c("Female", "RSLow:Female", "RSHigh:Female", "ESLow:Female", "ESHigh:Female")
power_lab   <- c("Female", "LR x Female", "HR x Female", "LE x Female", "HE x Female")

s2_power <- do.call(rbind, Map(function(t, l)
  power_row(l, cho_td[t, "b"], cho_td[t, "se"], N_s2), power_terms, power_lab))
rownames(s2_power) <- NULL
names(s2_power)[names(s2_power) == "Clustered SE"] <- "Bootstrap SE"

kable(s2_power, align = c("l", rep("r", 5)),
      caption = paste0("Study II - approximate post-estimation power diagnostics ",
        "for latent-index gender coefficients in the chosen-effort ordered-probit ",
        "model (N = ", N_s2, "). SEs are session-cluster bootstrap SEs. ",
        "Approximate diagnostics, not exact ex-ante calculations."))
```

---

# Exports

Tables and figures are written to `output/` for inclusion in the thesis.

```{r exports}
dir.create("output", showWarnings = FALSE)
if (file.exists("output/power_diagnostics.csv")) invisible(file.remove("output/power_diagnostics.csv"))
old_study2_files <- c("output/study2_regression_table.csv",
                      "output/study2_power.csv",
                      "output/study2_wald_chosen_effort.csv",
                      "output/wald_paragraphs.tex")
invisible(file.remove(old_study2_files[file.exists(old_study2_files)]))

## Tables (CSV).
write.csv(s1_table,             "output/study1_rd_table.csv", row.names = FALSE)
write.csv(s1_power,             "output/study1_power.csv",    row.names = FALSE)
write.csv(s2_table_full,        "output/study2_ordered_probit_table.csv",
          row.names = FALSE)
write.csv(wald_tbl,             "output/study2_ordered_probit_wald.csv",
          row.names = FALSE)
write.csv(s2_implied_effects,   "output/study2_ordered_probit_implied_effects.csv",
          row.names = FALSE)
write.csv(s2_predicted_effort,  "output/study2_ordered_probit_predicted_effort.csv",
          row.names = FALSE)
write.csv(s2_power,             "output/study2_ordered_probit_power.csv",
          row.names = FALSE)

## Wald-test paragraphs as ready-to-paste LaTeX.
writeLines(c("% Paragraph 1 (Section 5.2.2):", para1, "",
             "% Paragraph 2 (Section 5.2.3):", para2),
           "output/study2_ordered_probit_wald_paragraphs.tex")

## Figures (PNG).
ggsave("output/fig_study1_rd_male.png",   p_rd_male,   width = 7, height = 4.6, dpi = 300)
ggsave("output/fig_study1_rd_female.png", p_rd_female, width = 7, height = 4.6, dpi = 300)
ggsave("output/fig_study1_rd_full.png",   p_rd_full,   width = 7, height = 4.6, dpi = 300)
ggsave("output/fig_study2_chosen_effort.png", p_s2,    width = 8.5, height = 5.4, dpi = 300)

cat("Exported to ", normalizePath("output"), ":\n", sep = "")
cat(paste(" -", list.files("output")), sep = "\n")
```

```{r session-info}
sessionInfo()
```
